---
title: "Australian RSPCA "
output: html_notebook
---

```{r}
library(rpart)
library(rpart.plot)
library(here)
library(forcats)
library(janitor)
library(tidyverse)
```

The Data:

```{r}
townsville_complaints <- read_csv(here(
  "clean_data/townsville_complaints_clean.csv")
  )

townsville_complaints_tot_per_suburb <- read_csv(here(
  "clean_data/townsville_complaints_tot_per_suburb.csv")
  )

national_outcomes <- read_csv(here(
  "clean_data/national_outcomes_clean.csv")
  )

brisbane_complaints <- read_csv(here(
  "clean_data/brisbane_complaints_clean.csv")
  ) %>% 
  mutate(suburb = str_to_title(suburb)
         )

brisbane_complaints_tot_per_suburb <- read_csv(here(
  "clean_data/brisbane_complaints_tot_per_suburb.csv")
  ) %>% 
  mutate(suburb = str_to_title(suburb)
         )

townsville_complaints
townsville_complaints_tot_per_suburb
brisbane_complaints
brisbane_complaints_tot_per_suburb
national_outcomes
```
## Complains type plots for Townsville and Brisbane

```{r}
townsville_complaints_tot_plot <- townsville_complaints %>% 
  ggplot() +
  geom_bar(aes(y = fct_infreq(complaint_type), 
                                   fill = animal_type)) +
  labs(x = "Complaint Type", 
       y = "Totals", 
       title = "Townsville Complaint Type Totals") +
  scale_fill_manual("Animal Type", values = c("cat" = "#74c066",
                                              "dog" = "#0193cf")
                    ) +
  theme_bw() +
  theme(text=element_text(size = 16, family = "Arial"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  facet_wrap(~ animal_type)
  

townsville_complaints_tot_plot

ggsave("townsville_complaints_tot_plot.png")
```
In Townsville, dogs receive the more complaints.
Calls regarding Cats are more concentrated on the welfare of the animal rather than proper annoyance like in dogs 

```{r}
exclude <- c("Dog")

brisbane_complaints_tot_plot <- brisbane_complaints %>% 
  mutate(animal_type = str_to_lower(animal_type)) %>% 
  filter(animal_type %in% "cat" | animal_type %in% "dog") %>% 
  filter(!category %in% exclude) %>% 
  ggplot() +
  geom_bar(aes(y = fct_infreq(category), fill = animal_type)) +
  labs(x = "Complaint Type", 
       y = "Totals", 
       title = "Brisbane Complaint Type Totals") +
  scale_fill_manual("Animal Type", values = c("cat" = "#74c066",
                                              "dog" = "#0193cf")
                    ) +
  theme_bw() +
  theme(text=element_text(size = 16, family = "Arial"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  facet_wrap(~ animal_type)
  

brisbane_complaints_tot_plot

ggsave("brisbane_complaints_tot_plot.png")
```
```{r}
brisbane_complaints %>% 
  group_by(animal_type, category) %>% 
  summarise(n = n())
```


In Brisbane again, dogs are again the focus of complains.
We can notice some difference from Townsville, there is more variety of complains and looks like in Brisbane dogs have something for fences.
Also, it seems to be an issue with foxes. 
Let's look at this further.

## Brisbane vs Fox

```{r}
include <- "Fox"

brisbane_complaints_fox_plot <- brisbane_complaints %>%
  mutate(animal_type = str_to_lower(animal_type)) %>% 
 # filter(!animal_type %in% "cat" | animal_type %in% "dog") %>% 
  filter(category %in% include) %>% 
  ggplot() +
  geom_bar(position = "dodge", aes(x = category, fill = animal_type)) +
  labs(x = "Complaint Type", 
       y = "Totals", 
       title = "Brisbane Fox Troubles") +
  scale_fill_manual("Grievances", values = c("cat trapping" = "#73182c", 
                                             "other animal" = "#d6a115",
                                             "attack" = "#244321",
                                             "cat " = "#74c066",
                                             "dog" = "#0193cf")
                    ) +
  theme_bw() +
  theme(text=element_text(size = 16, family = "Arial"))
  
brisbane_complaints_fox_plot

ggsave("brisbane_complaints_fox_.png")
```
A fair amount of complaints are regarding foxes attacking dogs and cats and in particular trapping the latter.
- Advise business to do campaign on keeping pets inside at home as long as it necessary.

```{r}
brisbane_complaints_fox_per_suburbs <- brisbane_complaints_tot_per_suburb %>% 
  mutate(animal_type = str_to_lower(animal_type)) %>% 
  #filter(!animal_type %in% "cat" | animal_type %in% "dog") %>% 
  filter(category %in% include) %>% 
  filter(total > 30) %>% 
  ggplot() +
  geom_col(aes(y = suburb, x = total, fill = animal_type)) +
  labs(y = "Suburb", 
       x = "Totals", 
       title = "Brisbane Suburbs Fox Troubles*",
       caption = "* just considering Suburbs with more that 30 complaints") +
  scale_fill_manual("Grievances", values = c("cat trapping" = "#73182c",
                                              "attack" = "#244321",
                                              "other animal" = "#d6a115")
                    ) +
  scale_y_discrete(limits = rev) +
  theme_bw() +
  theme(text=element_text(size = 10, family = "Arial"),
        plot.caption = element_text(hjust = 1.4, face = "italic"))
  
brisbane_complaints_fox_per_suburbs

ggsave("brisbane_complaints_fox_per_suburbs.png")
```
If we look at the fox problem by suburb, we notice that the cat trapping issue is particularly felt in Inala. 

* insert map of Inala

```{r}
brisbane_complaints_by_suburbs <- brisbane_complaints_tot_per_suburb %>% 
  mutate(animal_type = str_to_lower(animal_type)) %>% 
  filter(animal_type %in% "cat" | animal_type %in% "dog") %>% 
  filter(!category %in% exclude) %>% 
  filter(total > 50) %>% 
  drop_na() %>% 
  ggplot() +
  geom_col(aes(y = suburb, x = total, fill = animal_type)) +
  labs(y = "Suburb", 
       x = "Totals", 
       title = "Brisbane Suburbs Pets Complaints*",
       caption = "* just considering Suburbs with more that 50 complaints") +
   scale_fill_manual("Animal Type", values = c("cat" = "#74c066",
                                              "dog" = "#0193cf")
                    ) +
  scale_y_discrete(limits = rev) +
  theme_bw() +
  theme(text=element_text(size = 10, family = "Arial"),
        axis.text.x = element_text(angle = 60, hjust = 1),
        plot.caption = element_text(hjust = 1.4, face = "italic"))
  
brisbane_complaints_by_suburbs

ggsave("brisbane_complaints_by_suburbs.png")
```
Looking at the total complaints by suburb that are not the conseguence of a fox, we can again notice the prevalence of Dogs over cats. 
Again Inala tops the other suburbs in complaints number.
In Bracken ridge there's seems to be too many stray cats.

- action in Bracken Ridge

```{r}
brisbane_subs_complaints_type_dogs <- brisbane_complaints_tot_per_suburb %>% 
  mutate(category = str_to_lower(category)) %>% 
  filter(animal_type == "Dog") %>% 
  filter(!category %in% exclude) %>% 
  filter(total > 50) %>% 
  drop_na() %>% 
  ggplot() +
  geom_col(aes(y = suburb, x = total, fill = category)) +
  labs(y = "Suburb", 
       x = "Totals", 
       title = "Brisbane Suburbs Pets Complaints Type - Dogs*",
       caption = "* just considering Suburbs with more that 50 complaints") +
  scale_fill_manual("Complaints Type", values = c("fencing issues" = "#74c066",
                                                  "unregistered" = "#244321",
                                                  "wandering" = "#d6a115")
                    ) +
  scale_y_discrete(limits = rev) +
  theme_bw() +
  theme(text=element_text(size = 10, family = "Arial"),
        plot.caption = element_text(hjust = 1.4, face = "italic"))

brisbane_subs_complaints_type_dogs

ggsave("brisbane_subs_complaints_type_dogs.png")
  
```
Looking deeper to the type of complaints regarding dogs, most of them are for fencing issue, troubles between neighbors.

- prevent dogs damage fence?
```{r}
brisbane_subs_complaints_type_cats <- brisbane_complaints_tot_per_suburb %>% 
  mutate(category = str_to_lower(category)) %>% 
  filter(animal_type == "Cat") %>% 
  filter(!category %in% exclude) %>% 
  filter(total > 10) %>% 
  drop_na() %>% 
  ggplot() +
  geom_col(aes(y = suburb, x = total, fill = category)) +
  labs(y = "Suburb", 
       x = "Totals", 
       title = "Brisbane Suburbs Pets Complaints Type - Cats*",
       caption = "* just considering Suburbs with more that 10 complaints") +
  scale_fill_manual("Complaints Type", values = c("fox" = "#74c066",
                                                  "wandering" = "#0193cf")
                    ) +
  scale_y_discrete(limits = rev) +
  theme_bw() +
  theme(text=element_text(size = 10, family = "Arial"),
        plot.caption = element_text(hjust = 1.4, face = "italic"))

brisbane_subs_complaints_type_cats

ggsave("brisbane_subs_complaints_type_cats.png")
```
If we look at cats, again the two main problems are wandering and foxes.

```{r}
brisbane_complaints %>% 
  filter(animal_type == "Dog" | animal_type == "Cat") %>% 
  group_by(animal_type, suburb) %>% 
  summarise(n = n()) %>% 
  arrange(desc(n))
```





```{r}
townsville_complaints_by_suburbs <- townsville_complaints_tot_per_suburb %>%
  filter(total > 30) %>% 
  ggplot() +
  geom_col(aes(y = suburb, x = total, fill = animal_type)) +
  labs(y = "Suburb", 
       x = "Totals", 
       title = "Townsville Suburbs Complaints Totals",
       caption = "* just considering Suburbs with more that 30 complaints") +
  scale_fill_manual("Complaints Type", values = c("cat" = "#74c066",
                                                  "dog" = "#0193cf")
                    ) +
  scale_y_discrete(limits = rev) +
  theme_bw() +
  theme(text=element_text(size = 10, family = "Arial"),
        plot.caption = element_text(hjust = 1.4, face = "italic"))

townsville_complaints_by_suburbs

ggsave("townsville_complaints_by_suburbs.png")
```
Going to Townsville suburbs, and dogs are at it again. Most complains are directed to dogs

```{r}
townsville_complaints_type_cats <- townsville_complaints_tot_per_suburb %>% 
  mutate(complaint_type = str_to_lower(complaint_type)) %>% 
  filter(animal_type == "cat") %>% 
  filter(total > 10) %>% 
  filter(!suburb == "Unallocated") %>% 
  ggplot() +
  geom_col(aes(y = suburb, x = total, fill = complaint_type)) +
  labs(y = "Suburb", 
       x = "Totals", 
       title = "Townsville Suburbs Complaints Cats",
       caption = "* just considering Suburbs with more that 10 complaints") +
  scale_fill_manual("Complaints Type", values = c("enclosure" = "#74c066",
                                                  "private impound" = "#0193cf",
                                                  "wandering" = "#244321")
                    ) +
  scale_y_discrete(limits = rev) +
  theme_bw() +
  theme(text=element_text(size = 10, family = "Arial"),
        axis.text.x = element_text(angle = 60, hjust = 1),
        plot.caption = element_text(hjust = 1.4, face = "italic"))

townsville_complaints_type_cats

ggsave("townsville_complaints_type_cats.png")
```
Again for cats we notice that most if not all the complaints calls are regarding the welfare of the animal, private impound means when a person than it's not part of RSCA contacts them because he took in an animal because they are concerned regarding its state. Enclosure is when an animal is trapped and wandering is moslty directed to strays or lost pets.

```{r}
townsville_complaints_type_dogs <- townsville_complaints_tot_per_suburb %>% 
  mutate(complaint_type = str_to_lower(complaint_type)) %>% 
  filter(!animal_type == "cat") %>% 
  filter(total > 50) %>% 
  filter(!suburb == "Unallocated") %>% 
  ggplot() +
  geom_col(aes(y = suburb, x = total, fill = complaint_type)) +
  labs(y = "Suburb", 
       x = "Totals", 
       title = "Townsville Suburbs Complaints Totals",
       caption = "* just considering Suburbs with more that 50 complaints") +
  scale_fill_manual("Complaints Type", values = c("enclosure" = "#74c066",
                                                  "private impound" = "#0193cf",
                                                  "wandering" = "#244321",
                                                  "attack" = "#d6a115",
                                                  "noise" = "#73182c",
                                                  "aggressive animal" = "#ff0000")
                    ) +
  scale_y_discrete(limits = rev) +
  theme_bw() +
  theme(text=element_text(size = 10, 0.1, family = "Arial"),
        plot.caption = element_text(hjust = 1.4, face = "italic")) 

townsville_complaints_type_dogs

ggsave("townsville_complaints_type_dogs.png")
```
Mount Louisa and Kirwan have a very high volume of calls for aggressive dogs and even attacks. There is a high level of call for enclosure as well, which could be either a trapped animal or most probably in this case a dog tied to a chain outside in hot water or closed inside without the opportunity to go out, this again reflects in the high than average amount of calls for noise.

RSPCA should probably led a further enquire on pets welfare in this suburbs and concentrate a campaign on how to be a good dog owner.



```{r}
townsville_dog_bars <- townsville_complaints %>% 
  filter(animal_type == "dog") %>%
  group_by(complaint_type) %>% 
  summarise(total = n()) %>% 
  arrange(desc(total)) 

townsville_dog_bars

ggplot(townsville_dog_bars) +
  geom_col(aes(x = complaint_type, y = total))

  

```

```{r}
townsville_cat_bars <- townsville_complaints %>% 
  filter(animal_type == "cat") %>%
  group_by(complaint_type) %>% 
  summarise(total = n()) %>% 
  arrange(desc(total)) 

townsville_cat_bars

ggplot(townsville_cat_bars) +
  geom_col(aes(x = complaint_type, y = total))
```


```{r}
brisbane_dog_bars <- brisbane_complaints %>% 
  filter(animal_type == "Dog") %>%
  group_by(category) %>% 
  summarise(total = n()) %>% 
  arrange(desc(total)) 

brisbane_dog_bars

ggplot(brisbane_dog_bars) +
  geom_col(aes(x = category, y = total))

  

```

```{r}
brisbane_cat_bars <- brisbane_complaints %>% 
  filter(animal_type == "Cat") %>%
  group_by(category) %>% 
  summarise(total = n()) %>% 
  arrange(desc(total)) 

brisbane_cat_bars

ggplot(brisbane_cat_bars) +
  geom_col(aes(x = category, y = total))

```

```{r}
national_outcomes_animail_type <-  national_outcomes %>% 
  filter(!outcome == "Currently In Care") %>% # Currently in Care refers to animal in RSPCA custody waiting for an outcome, as the outcome is not defined yet for these cases I have excluded it from the plot
  mutate(animal_type = str_to_lower(animal_type)) %>% 
  group_by(animal_type, outcome) %>% 
  summarise(tot = sum(tot_outcomes)) %>% 
  drop_na() %>% 
  ggplot() +
  geom_col(position = "dodge", colour = "white", aes(x = outcome, 
                                                     y = tot, 
                                                     fill = animal_type)
           )+
  labs(x = "Outcome", 
       y = "Totals", 
       title = "National Complaint Outcome per animal type since 1999") +
  scale_fill_manual("Animal Type", values = c("cats" = "#74c066",
                                                  "dogs" = "#0193cf",
                                                  "horses" = "#244321",
                                                  "livestock" = "#d6a115",
                                                  "other animals" = "#73182c",
                                                  "wildlife" = "#ff0000")
                    ) +
  scale_y_continuous(labels = scales::comma) +
  theme_bw() +
  theme(text=element_text(size = 14, family = "Arial"),
        axis.text.x = element_text(angle = 45, hjust = 1))

national_outcomes_animail_type  
```
```{r}
national_outcomes_pets <-  national_outcomes %>% 
  filter(!outcome == "Currently In Care") %>%
  filter(!outcome == "Other") %>%
  filter(animal_type == "Cats" | animal_type == "Dogs") %>%
  mutate(animal_type = str_to_lower(animal_type)) %>% 
  group_by(animal_type, outcome) %>% 
  summarise(tot = sum(tot_outcomes)) %>% 
  drop_na() %>% 
  ggplot() +
  geom_col(position = "dodge", colour = "white", aes(y = fct_infreq(outcome), 
                                                     x = tot, 
                                                     fill = animal_type)
           )+
  labs(y = "Outcome", 
       x = "Totals", 
       title = "National Complaint Outcome per animal type since 1999") +
  scale_fill_manual("Animal Type", values = c("cats" = "#74c066",
                                                  "dogs" = "#0193cf")
                    ) +
  scale_x_continuous(labels = scales::comma) +
  theme_bw() +
  theme(text=element_text(size = 14, family = "Arial"),
        axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none") +
  facet_wrap(~animal_type)

national_outcomes_pets

ggsave("national_outcomes_pets.png")

```

The outcome for complaint when the animal is taken in custody by RSPCA from 1999.
The most common outcome for cats is to be euthanised, and they are not reclaimed as often a dogs and I am not sure how many how the rehoming are successful, I can imagine that a cat that gets returned to RSPCA or ends up again in the street is not a rare scenario, that would explain the difference in numbers for euthanised and reclaimed with dogs.

The number of cats and dogs in the data set is almost equivalent so I tend to belive the above hypothesis.

Cats	1117506			
Dogs	1149320

```{r}
euthanisation_trend_cats_dogs <- national_outcomes %>%  
  filter(animal_type == "Cats" | animal_type == "Dogs") %>% 
  filter(outcome == "Euthanized") %>%
  mutate(animal_type = str_to_lower(animal_type)) %>% 
  group_by(year, animal_type, outcome) %>% 
  summarise(tot_outcomes = sum(tot_outcomes)) %>% 
  ungroup() %>% 
  ggplot() +
  geom_line(size = 2,aes(x = year, 
                         y = tot_outcomes, 
                         group = animal_type, 
                         colour = animal_type)
            ) +
  labs(x = "Year", 
       y = "Euthanisations", 
       title = "Cats & Dogs Euthanisation trend since 1999") +
  scale_colour_manual("Animal Type", values = c("cats" = "#74c066",
                                              "dogs" = "#0193cf")
                    ) +
  theme_bw() +
  theme(text=element_text(size = 14, family = "Arial"))

euthanisation_trend_cats_dogs

ggsave("euthanisation_trend_cats_dogs.png")
```
The good news is that the trend is changing, since the 2010s euthanisations have dropped drastically.

```{r}
rehoming_trendcats_dogs <- national_outcomes %>%  
  filter(animal_type == "Cats" | animal_type == "Dogs") %>% 
  filter(outcome == "Rehomed") %>%
  mutate(animal_type = str_to_lower(animal_type)) %>% 
  group_by(year, animal_type, outcome) %>% 
  summarise(tot_outcomes = sum(tot_outcomes)) %>% 
  ungroup() %>% 
  ggplot() +
  geom_line(size = 2,aes(x = year, 
                         y = tot_outcomes, 
                         group = animal_type, 
                         colour = animal_type)
            ) +
  labs(x = "Year", 
       y = "Rehomed", 
       title = "Cats & Dogs Rehomed trend since 1999") +
  scale_colour_manual("Animal Type", values = c("cats" = "#74c066",
                                              "dogs" = "#0193cf")
                    ) +
  theme_bw() +
  theme(text=element_text(size = 14, family = "Arial"))

rehoming_trendcats_dogs

ggsave("rehoming_trendcats_dogs.png")
```
Rehoming numbers for cats have increase drastically since the 2010s, but again I have doubts on how many of these were succesful.

```{r}
reclaimed_trendcats_dogs <- national_outcomes %>%  
  filter(animal_type == "Cats" | animal_type == "Dogs") %>% 
  filter(outcome == "Reclaimed") %>%
  mutate(animal_type = str_to_lower(animal_type)) %>% 
  group_by(year, animal_type, outcome) %>% 
  summarise(tot_outcomes = sum(tot_outcomes)) %>% 
  ungroup() %>% 
  ggplot() +
  geom_line(size = 2,aes(x = year, 
                         y = tot_outcomes, 
                         group = animal_type, 
                         colour = animal_type)
            ) +
  labs(x = "Year", 
       y = "Recaimed", 
       title = "Cats & Dogs Reclaimed trend since 1999") +
  scale_colour_manual("Animal Type", values = c("cats" = "#74c066",
                                              "dogs" = "#0193cf")
                    ) +
  theme_bw() +
  theme(text=element_text(size = 14, family = "Arial"))

reclaimed_trendcats_dogs

ggsave("reclaimed_trendcats_dogs.png")
```
The reclaimed trends for cats hasn't been that great and it's stable during the years.
For dog the trend is going down although this couldn't been so gloom, with the diffusion of social media and in particular neighbourhood pages, when a dog go missing the voice is spread quickly and usualy they are found before RSPCA intervention. 

```{r}
national_outcomes %>% 
  filter(animal_type == "Cats" | animal_type == "Dogs") %>% 
  group_by(animal_type) %>% 
  drop_na() %>% 
  summarise(tot_outcomes = sum(tot_outcomes))
```

```{r}
national_outcomes %>% 
  filter(animal_type == "Dogs" | animal_type == "Cats") %>% 
  filter(outcome == "Euthanized" | 
         outcome == "Rehomed" | 
         outcome == "Reclaimed") %>% 
  group_by(animal_type, outcome) %>% 
  drop_na() %>% 
  summarise(tot = sum(tot_outcomes))
```

Plain numbers above

## Decision Tree


Preparing the data set for the decision tree

```{r}
tree_national_outcomes <- rowid_to_column(national_outcomes, "id") %>% 
  drop_na()

decision_tree_try <- data.frame(lapply(tree_national_outcomes, 
                                       rep, 
                                       tree_national_outcomes$tot_outcomes))
```

```{r}
clean_national_test <- decision_tree_try %>% 
  #filter(year == "2018") %>% 
  filter(animal_type == "Dogs" | animal_type == "Cats") %>% 
  filter(outcome == "Euthanized" | 
         outcome == "Rehomed" ) %>% 
   mutate(animal_type = as.factor(animal_type),
         state = as.numeric(factor(state)),
         outcome = as.numeric(factor(outcome))
         ) %>% 
  mutate(state = factor(state, levels = c(1,2,3,4,5,6,7,8),
                        labels = c("Australian Capital Territory",
                                   "New South Wales",
                                   "Northern Territory",
                                   "Queensland",
                                   "South Australia",
                                   "Tasmania",
                                   "Victoria",
                                   "Western Australia")),
         outcome = factor(outcome, levels = c(1,2), 
                              labels = c("Euthanized", "Rehomed"))
         ) %>% 
  select(-id, -tot_outcomes) 

clean_national_test
```

```{r}
n_data <- nrow(clean_national_test)

test_index <- sample(1:n_data, size = n_data*0.2)

outcomes_test  <- slice(clean_national_test, test_index)

outcomes_train <- slice(clean_national_test, -test_index)

outcomes_test %>% 
  tabyl(outcome)

outcomes_train %>% 
  tabyl(outcome)

```

```{r}
outcome_fit_state <- rpart(
  formula = outcome ~ animal_type + state, 
  data = outcomes_train, 
  method = 'class'
)

rpart.plot(outcome_fit_state , 
           yesno = 2,
           fallen.leaves = TRUE, 
           faclen = 6, 
           digits = 2)
```
		


```{r}
outcome_fit_animal <- rpart(
  formula = outcome ~ animal_type, 
  data = outcomes_train, 
  method = 'class'
)

rpart.plot(outcome_fit_animal, 
           yesno = 2,
           fallen.leaves = TRUE, 
           faclen = 6, 
           digits = 2)
```

```{r}
outcome_fit_all <- rpart(
  formula = outcome ~ ., 
  data = outcomes_train, 
  method = 'class'
)

rpart.plot(outcome_fit_all, 
           yesno = 2,
           fallen.leaves = TRUE, 
           faclen = 6, 
           digits = 2)
```





